Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-04-29 doc/paper.Rmd
almost done with discussion

-   @criley1994 : under hort conditions, shoots that flower in spring emerge in June/July the previous year.
    \~6 months after shoot emergence that the developing infloresence is susceptible to adverse effects.

@@

-   @westerbandEarlyLifeConditions2017: *Heliconia* grew more at low than high precipitation.
    Leaf life span and biomass allocation did not differ among treatments.
    The effect of environmental variability was dependent on early life conditions as well as precipitation, suggesting that generalist species may experience high fitness as forest environments become more variable by maintaining high growth at the expense of survival.

@@

-   @wrightSeasonalDroughtSoil1992: seasonal drought.
    experimental irrigation influences survival and growth of tropical herbs

@@

    -   low precip = trees lose leaves or die, more light gets through or gaps open.
        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020)

-   @horvitzDemographicCostReproduction1988: no evidence for short term (next season) cost of reproduction in another understory monocot (Calathea) BUT

-   @horvitzEffectsPlantSize2002 (*Calathea ovandensis*) Survival of juveniles was positively affected by plant size, but not by herbivory, local competition, or fruit production.
    Survival of large stages was not generally affected by any of the factors.
    Relative growth rates of all stages were negatively affected by plant size; larger plants grew more slowly than smaller plants.
    Additionally, growth rate of juveniles was negatively affected by local competition, whereas growth rate of reproductives was positively affected by reproduction.
    Next year's inflorescence production was positively affected by plant size for both pre-reproductive and reproductive stages.
    For reproductives, it was additionally positively affected by current fruit production.

@@

-   @stiles1975?
    drought resulted in less flowering in Heliconia

-   @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet

@@

Precipitation and other climatic factors are known to have large effects on plant growth, survival, or reproduction.
Despite this, models of population dynamics have rarely included the effects of environmental drivers.
This has largely due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021]. Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

However, while it appeared that delayed effects of climate on demographic vital rates were ubiquitous, the extent to which they vary spatially or with habitat remained an open question [@eversLaggedDormantSeason2021]. Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates. For example, [delete this phrase? the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than frragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for the effects of SPEI on vital rates could take months or even years to manifest.
First, these delayed effects may represent a physiological delayed effect of environmental conditions.
Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@crileyYearProductionHigh1994]. <!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H. acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.  "Differentiation of the apical meristem" ? --> likely happens many months earlier. The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation. e.g. S. Joe Wright trees flowering after ENSO -->.
We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
Our models show that drought two dry seasons prior to the census increase flowering probability.
A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.

Second, delayed effects could be mediated by indirect effects and ecological interactions.
<!--# I'm not sure if this is really an explanation for the *delay* per se.  Need to think about this. -->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration to the understory (<!--# citations -->).
This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
<!--# Heliconia grow better at edges, where there's more light --> Janssen et al. [@janssenDroughtEffectsLeaf2021] showed that increase in litterfall due to drought was delayed.
Here, the delayed effect of drought on *H. acuminata* could be partly explained by a delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.
Furthermore, literfall in response to drought has shown to be greater at forest edges compared to interiors, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments.
Drought is also associated with increased insolation, so negative SPEI could also be a proxy for increased insolation, which would be exacerbated at forest edges potentially.

Finally, it could be an artefact of the timing of responses vis-à-vis the census date.
If extreme drought in the dry season before the census increases plant mortality in that season, for example, this would show up as a delayed effect in our models (e.g. in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
This explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.

@@

-

>>>>

@@

@@

@@

    -   low precip = trees lose leaves or die, more light gets through or gaps open.
        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20

@@

    @mulkeyInfluenceSeasonalDrought1996 and @bonalResponseTropicalRainforests2016 will have goods stuff; haven't looked them over yet

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including tropical understory herbs [@wrightSeasonalDroughtSoil1992]. For instance, @westerbandEarlyLifeConditions2017: found that *Heliconia* grew more at low than high precipitation, while @stiles1975: observed reductions in flowering as a result of drought. Despite this, models of population dynamics have rarely included the effects of environmental drivers. This has largely been due to the challenge - both ecologically and statistically - of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons. To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature -- that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to **XXX growing seasons**.

While it appears that delayed effects of climate on demographic vital rates are ubiquitous, however, the extent to which they vary spatially or with habitat remains an open question [@eversLaggedDormantSeason2021]. Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates. For example, [delete this phrase? the overall survival of plants was similar in continuous forest and forest fragments, and] although extreme values of SPEI - both positive (unusually high precipitation) and negative (drought conditions) - led to declines in the probability of individual survival both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects on growth rates of variation in SPEI, in this case unusually rainy wet seasons and dry seasons with unusually low precipitation, were also more severe in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically relevant.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 prior to the appearance of infloresences. Adverse conditions during 6 months following initiation, rather than the months when infloresences are starting expand, is what leads to the aborted reproduction of flowering shoots [@crileyYearProductionHigh1994]. Interestingly, we observed the opposite effect - drought two dry seasons prior to a census actually increased the observed probability of flowering. While this could reflect bet-hedging in response to stress [@nihad2018], this does not appear to be the case, and there do not appear to be subsequent decreases in growth or survival following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering is associated with increased reproduction and growth in the following year **(is this an acceptable colloquial way to interpret the covariate effect?)**.

Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
For example, the mortality rate of tropical trees can spike in the year following a severe drought.
The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments (**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?**)**.
Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon). Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.

Finally, demographic delays could be an artefact of the timing of responses vis-à-vis the census date.
If extreme drought in the dry season before the census increased plant mortality during that season, for example, this would nevertheless appear in models as a delayed effect (e.g. in Figure \@ref(fig:surv-spei)b).
In our case, this potential explanation for delayed effects applies only to plant size and survival, as plots were surveyed regularly throughout the reproductive season to identify flowering plants.
[consequence of demography on an annual scale and weather on a monthly / seasonal scale.
This problem not unique to us, same for everyone.
Mismatched scale at which analyses operate]

@@

-   More experiments like westerband that try and see how contiions today influence growth in future years Generalizations about the direction and magnitude of delayed demographic responses will continue to prove elusive.
    Without experiments designed greater understanding of the complex ways in which light, environmental conditions, and other ecological conditions interactions to influence plant vital rates (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002),
